---
title: "Tutorial 1"
author: "Lauren Blake"
date: 2018-04-11
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

# Introduction

The goal of this script is to install the WGCNA package and then to explore [tutorial 1](https://labs.genetics.ucla.edu/horvath/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/index.html). According to the website, "the first tutorial guides the reader through an analysis of a single empirical gene expression data set."

Note: Most of the code below was taken from the R scripts also provided by the WGCNA authors and can be found on the tutorial [website](https://labs.genetics.ucla.edu/horvath/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/index.html). Most of the annotation outside of the code chunks is my own. 

Note: The networks, modules, etc. below were build using parameters given in the tutorial. I will need to spend time learning about these parameters and exploring the robustness of these parameters. 

# Data input and cleaning

**This includes re-formatting the data for analysis with the gene expression levels and the traits.** 

```{r}
# Display the current working directory
getwd()

# Load the package
library(WGCNA)
library(AnnotationDbi)
library(anRichment)

# The following setting is important, do not omit.
options(stringsAsFactors = FALSE)
enableWGCNAThreads()
allowWGCNAThreads()
#Read in the female liver data set
femData = read.csv("../data/LiverFemale3600.csv")
# Read in the male liver data set
maleData = read.csv("../data/LiverMale3600.csv");
# Take a quick look at what is in the data sets:
dim(femData)
#names(femData)
dim(maleData)
#names(maleData)

# Reshape the data so that we have 1 gene/microarray measurement per column and each sample name per row. 
datExpr0 = as.data.frame(t(femData[, -c(1:8)]));
dim(datExpr0)
names(datExpr0) = femData$substanceBXH;
rownames(datExpr0) = names(femData)[-c(1:8)];

```

**Here, we have microarray data from 135 female mouse liver samples and 3600 genes/microarray positions.** 

## Filter samples and genes with too many missing entries

```{r}
gsg = goodSamplesGenes(datExpr0, verbose = 3);
gsg$allOK

if (!gsg$allOK)
{
  # Optionally, print the gene and sample names that were removed:
  if (sum(!gsg$goodGenes)>0) 
     printFlush(paste("Removing genes:", paste(names(datExpr0)[!gsg$goodGenes], collapse = ", ")));
  if (sum(!gsg$goodSamples)>0) 
     printFlush(paste("Removing samples:", paste(rownames(datExpr0)[!gsg$goodSamples], collapse = ", ")));
  # Remove the offending genes and samples from the data:
  datExpr0 = datExpr0[gsg$goodSamples, gsg$goodGenes]
}

```
**All of uur samples passed this filtering step.**

# Sample clustering

This clustering is based on gene expression data for each sample. 

```{r}
sampleTree = hclust(dist(datExpr0), method = "average");
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5, 
     cex.axis = 1.5, cex.main = 2)

# Plot a line to show the cut
abline(h = 15, col = "red")
# Determine cluster under the line
clust = cutreeStatic(sampleTree, cutHeight = 15, minSize = 10)
table(clust)
# clust 1 contains the samples we want to keep.
keepSamples = (clust==1)
datExpr = datExpr0[keepSamples, ]
nGenes = ncol(datExpr)
nSamples = nrow(datExpr)
```
**134/135 samples do not have evidence of being outliers.**

# Add the clinical data

```{r}
traitData = read.csv("../data/ClinicalTraits.csv");
dim(traitData)


# remove columns that hold information we do not need.
allTraits = traitData[, -c(31, 16)];
allTraits = allTraits[, c(2, 11:36) ];
dim(allTraits)


# Form a data frame analogous to expression data that will hold the clinical traits.

femaleSamples = rownames(datExpr);
traitRows = match(femaleSamples, allTraits$Mice);
datTraits = allTraits[traitRows, -1];
rownames(datTraits) = allTraits[traitRows, 1];

dim(datTraits)

names(datTraits)

```
**There are 25 traits, many of which are related to weight and metabolism.**

## Cluster samples with gene expression and make heatmap of the clinical traits

```{r}
# Re-cluster samples
sampleTree2 = hclust(dist(datExpr), method = "average")
# Convert traits to a color representation: white means low, red means high, grey means missing entry
traitColors = numbers2colors(datTraits, signed = FALSE);
# Plot the sample dendrogram and the colors underneath.
plotDendroAndColors(sampleTree2, traitColors,
                    groupLabels = names(datTraits), 
                    main = "Sample dendrogram and trait heatmap")

save(datExpr, datTraits, file = "../data/FemaleLiver-01-dataInput.RData")
```

**Legend: white means low, red means high, grey means missing entry**

# Gene expression network construction and module detection

```{r}
# Load the data saved in the previous section
lnames = load(file = "../data/FemaleLiver-01-dataInput.RData")

# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to=20, by=2))
# Call the network topology analysis function
sft = pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)
# Plot the results:
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers,col="red")

softPower = 6;
adjacency = adjacency(datExpr, power = softPower)

# Turn adjacency into topological overlap
TOM = TOMsimilarity(adjacency);
dissTOM = 1-TOM
```

## Hierarchical clustering with TOM (Topological Overlap Matrix)-based dissimilarity

```{r}
geneTree = hclust(as.dist(dissTOM), method = "average");
# Plot the resulting clustering tree (dendrogram)
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
     labels = FALSE, hang = 0.04)

```

## Partion genes with similar gene expression into modules

```{r}
# We like large modules, so we set the minimum module size relatively high:
minModuleSize = 30;
# Module identification using dynamic tree cut:
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM,
                deepSplit = 2, pamRespectsDendro = FALSE,
                minClusterSize = minModuleSize);
table(dynamicMods)

# Give each module a color

# Convert numeric lables into colors
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "Gene dendrogram and module colors")

```

**Each color represents a different module.**

## Cluster and merge modules based on eigengenes

```{r}
# Calculate eigengenes
MEList = moduleEigengenes(datExpr, colors = dynamicColors)
MEs = MEList$eigengenes
# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs);
# Cluster module eigengenes
METree = hclust(as.dist(MEDiss), method = "average");
# Plot the result
plot(METree, main = "Clustering of module eigengenes",
     xlab = "", sub = "")
MEDissThres = 0.25
# Plot the cut line into the dendrogram
abline(h=MEDissThres, col = "red")
# Call an automatic merging function
merge = mergeCloseModules(datExpr, dynamicColors, cutHeight = MEDissThres, verbose = 3)
# The merged module colors
mergedColors = merge$colors;
# Eigengenes of the new merged modules:
mergedMEs = merge$newMEs;

#pdf(file = "Plots/geneDendro-3.pdf", wi = 9, he = 6)
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors),
                    c("Dynamic Tree Cut", "Merged dynamic"),
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
#dev.off()

# Rename to moduleColors
moduleColors = mergedColors
# Construct numerical labels corresponding to the colors
colorOrder = c("grey", standardColors(50));
moduleLabels = match(moduleColors, colorOrder)-1;
MEs = mergedMEs;
# Save module colors and labels for use in subsequent parts
save(MEs, moduleLabels, moduleColors, geneTree, file = "../data/FemaleLiver-02-networkConstruction-stepByStep.RData")

```

**Each color represents a different merged module.**

# Begin to study relationships between traits and modules

```{r}
# Load the expression and trait data saved in the first part
lnames = load(file = "../data/FemaleLiver-01-dataInput.RData");
#The variable lnames contains the names of loaded variables.
lnames
# Load network data saved in the second part.
lnames = load(file = "../data/FemaleLiver-02-networkConstruction-stepByStep.RData");
lnames

# Define numbers of genes and samples
nGenes = ncol(datExpr);
nSamples = nrow(datExpr);
# Recalculate MEs with color labels
MEs0 = moduleEigengenes(datExpr, moduleColors)$eigengenes
MEs = orderMEs(MEs0)
moduleTraitCor = cor(MEs, datTraits, use = "p");
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);


# Will display correlations. For space reasons, we will not display the p values of the correlations. 
textMatrix =  signif(moduleTraitCor, 2)
textMatrix = formatC(moduleTraitCor, digits = 2, format = "f")

dim(textMatrix) = dim(moduleTraitCor)

# Display the correlation values within a heatmap plot
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = names(datTraits),
               yLabels = names(MEs),
               ySymbols = NULL,
               colorLabels = FALSE,
               colors = blueWhiteRed(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(-1,1), 
               main = paste("Gene Expression Module- Trait relationships"), ylab = "Gene Expression-Based Modules")
```


# Begin to explore key gene drivers in interesting modules 

In this example, we will identify the key genes for weight in the brown module. 
```{r}
# Define variable weight containing the weight column of datTrait
weight = as.data.frame(datTraits$weight_g)
names(weight) = "weight"
# names (colors) of the modules
modNames = substring(names(MEs), 3)

geneModuleMembership = as.data.frame(cor(datExpr, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

geneTraitSignificance = as.data.frame(cor(datExpr, weight, use = "p"));
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance), nSamples));

names(geneTraitSignificance) = paste("GS.", names(weight), sep="");
names(GSPvalue) = paste("p.GS.", names(weight), sep="")

module = "brown"
column = match(module, modNames);
moduleGenes = moduleColors==module;



# Identify genes in the brown module
#names(datExpr)[moduleColors=="brown"]
```
These are all genes that were assigned to the brown module (based on gene expression values across the mice samples). 

## Identify genes in the brown module that are significant for body weight
```{r}

verboseScatterplot(abs(geneModuleMembership[moduleGenes, column]),
                   abs(geneTraitSignificance[moduleGenes, 1]),
                   xlab = paste("Module Membership in", module, "module"),
                   ylab = "Gene significance for body weight",
                   main = paste("Module membership vs. gene significance\n"),
                   cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2, col = module)



```

```{r}
module = "blue"
column = match(module, modNames);
moduleGenes = moduleColors==module;


verboseScatterplot(abs(geneModuleMembership[moduleGenes, column]),
                   abs(geneTraitSignificance[moduleGenes, 1]),
                   xlab = paste("Module Membership in", module, "module"),
                   ylab = "Gene significance for body weight",
                   main = paste("Module membership vs. gene significance\n"),
                   cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2, col = module)
```


```{r}
module = "black"
column = match(module, modNames);
moduleGenes = moduleColors==module;


verboseScatterplot(abs(geneModuleMembership[moduleGenes, column]),
                   abs(geneTraitSignificance[moduleGenes, 1]),
                   xlab = paste("Module Membership in", module, "module"),
                   ylab = "Gene significance for body weight",
                   main = paste("Module membership vs. gene significance\n"),
                   cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2, col = module)

```

**Overall, the genes in different modules have different relationships to body weight. In future tutorials, I will extract the modules/genes in each module that are most related to each trait of interest.**



# Integrate the network analysis with functional enrichment 

**Begin by annotating the genes in the brown module.**

```{r}
module = "brown"
column = match(module, modNames);
moduleGenes = moduleColors==module;

annot = read.csv(file = "../data/GeneAnnotation.csv");
dim(annot)
names(annot)
probes = names(datExpr)
probes2annot = match(probes, annot$substanceBXH)
# The following is the number or probes without annotation:
sum(is.na(probes2annot))
# Should return 0.

# Create the starting data frame
geneInfo0 = data.frame(substanceBXH = probes,
                      geneSymbol = annot$gene_symbol[probes2annot],
                      LocusLinkID = annot$LocusLinkID[probes2annot],
                      moduleColor = moduleColors,
                      geneTraitSignificance,
                      GSPvalue)
# Order modules by their significance for weight
modOrder = order(-abs(cor(MEs, weight, use = "p")));
# Add module membership information in the chosen order
for (mod in 1:ncol(geneModuleMembership))
{
  oldNames = names(geneInfo0)
  geneInfo0 = data.frame(geneInfo0, geneModuleMembership[, modOrder[mod]], 
                         MMPvalue[, modOrder[mod]]);
  names(geneInfo0) = c(oldNames, paste("MM.", modNames[modOrder[mod]], sep=""),
                       paste("p.MM.", modNames[modOrder[mod]], sep=""))
}
# Order the genes in the geneInfo variable first by module color, then by geneTraitSignificance
geneOrder = order(geneInfo0$moduleColor, -abs(geneInfo0$GS.weight));
geneInfo = geneInfo0[geneOrder, ]

head(geneInfo)

# Save the file
write.csv(geneInfo, file = "../data/geneInfo.csv")
```

## Perform GO enrichment for each module

**Note: Since the function GoEnrichmentAnalysis is defunction, per the author's suggestion, we are using the function enrichmentAnalysis from the package "AnRichment" from the same authors. For more information about this package, go [here](https://labs.genetics.ucla.edu/horvath/htdocs/CoexpressionNetwork/GeneAnnotation/Tutorials/anRichment-Tutorial1.pdf).** 

```{r}
# Load the expression and trait data saved in the first part
lnames = load(file = "../data/FemaleLiver-01-dataInput.RData")
#The variable lnames contains the names of loaded variables.
lnames
# Load network data saved in the second part.
lnames = load(file = "../data/FemaleLiver-02-networkConstruction-stepByStep.RData")
lnames

# Read in the probe annotation
annot = read.csv(file = "../data/GeneAnnotation.csv");
# Match probes in the data set to the probe IDs in the annotation file 
probes = names(datExpr)
probes2annot = match(probes, annot$substanceBXH)
# Get the corresponding Locuis Link IDs
allLLIDs = annot$LocusLinkID[probes2annot];
# $ Choose interesting modules
intModules = c("brown", "red", "salmon")
for (module in intModules)
{
  # Select module probes
  modGenes = (moduleColors==module)
  # Get their entrez ID codes
  modLLIDs = allLLIDs[modGenes];
  # Write them into a file
  fileName = paste("LocusLinkIDs-", module, ".txt", sep="");
  write.table(as.data.frame(modLLIDs), file = fileName,
              row.names = FALSE, col.names = FALSE)
}
# As background in the enrichment analysis, we will use all probes in the analysis.
fileName = paste("../data/LocusLinkIDs-all.txt", sep="");
write.table(as.data.frame(allLLIDs), file = fileName,
            row.names = FALSE, col.names = FALSE)
```


```{r}
# Use the GO collection for mouse GO terms. 
GOcollection = buildGOcollection(organism = "mouse")

# Run the GO enrichment

GOenr = enrichmentAnalysis(
     classLabels = moduleColors, identifiers = allLLIDs,
     refCollection = GOcollection,
     useBackground = "allOrgGenes",
     threshold = 1e-4,
     thresholdType = "Bonferroni",
     getOverlapEntrez = TRUE,
     getOverlapSymbols = TRUE,
     ignoreLabels = "grey")


names(GOenr)
names(GOenr$dataSetDetails)
GOenr$dataSetDetails$black

# Make set with all enriched GO terms

GO_per_set <- cbind(GOenr$enrichmentTable$class, GOenr$enrichmentTable$dataSetID,  GOenr$enrichmentTable$dataSetName, GOenr$enrichmentTable$FDR)

tail(GO_per_set)
colnames(GO_per_set) <- c("Module", "GO term", "GO process", "FDR")

# Pull the top (most significant) for each module

top_GO_per_module <- GO_per_set[!duplicated(GO_per_set[,1]), ]
colnames(top_GO_per_module) <- c("Module", "GO term", "GO process", "FDR")
```
**As expected, the most enriched GO terms for each module are extremely general (e.g. "cell part"). As we decrease in significance, we get more specific/likely informative ones (e.g. "positive regulation of protein modification process").**

# Additional visualizations

**These are additional plots recommended by the WGCNA authors. I am working to get information that will make these plots more intuitive.**
```{r}
nGenes = ncol(datExpr)
nSamples = nrow(datExpr)


# Calculate topological overlap anew: this could be done more efficiently by saving the TOM
# calculated during module detection, but let us do it again here.
dissTOM = 1-TOMsimilarityFromExpr(datExpr, power = 6);
# Transform dissTOM with a power to make moderately strong connections more visible in the heatmap
plotTOM = dissTOM^7;
# Set diagonal to NA for a nicer plot
diag(plotTOM) = NA;
# Call the plot function
sizeGrWindow(9,9)
TOMplot(plotTOM, geneTree, moduleColors, main = "Network heatmap plot, all genes")
```


```{r}
nSelect = 400
# For reproducibility, we set the random seed
set.seed(10);
select = sample(nGenes, size = nSelect);
selectTOM = dissTOM[select, select];
# There's no simple way of restricting a clustering tree to a subset of genes, so we must re-cluster.
selectTree = hclust(as.dist(selectTOM), method = "average")
selectColors = moduleColors[select];
# Open a graphical window
sizeGrWindow(9,9)
# Taking the dissimilarity to a power, say 10, makes the plot more informative by effectively changing 
# the color palette; setting the diagonal to NA also improves the clarity of the plot
plotDiss = selectTOM^7;
diag(plotDiss) = NA;
TOMplot(plotDiss, selectTree, selectColors, main = "Network heatmap plot, selected genes")

```

## Which gene co-expression network best explains one of the traits, weight?

**Based on the clustering below, it is co-expression network represented by the color green.**
```{r}
# Additional visualizations

# Recalculate module eigengenes
MEs = moduleEigengenes(datExpr, moduleColors)$eigengenes
# Isolate weight from the clinical traits
weight = as.data.frame(datTraits$weight_g);
names(weight) = "weight"
# Add the weight to existing module eigengenes
MET = orderMEs(cbind(MEs, weight))
# Plot the relationships among the eigengenes and the trait
#sizeGrWindow(5,7.5);
#par(cex = 0.9)
plotEigengeneNetworks(MET, "", marDendro = c(0,4,1,2), marHeatmap = c(3,4,1,2), cex.lab = 0.8, xLabelsAngle
= 90)

# Plot the dendrogram
#sizeGrWindow(6,6);
#par(cex = 1.0)
plotEigengeneNetworks(MET, "Eigengene dendrogram", marDendro = c(0,4,2,0),
                      plotHeatmaps = FALSE)
# Plot the heatmap matrix (note: this plot will overwrite the dendrogram plot)
#par(cex = 1.0)
plotEigengeneNetworks(MET, "Eigengene adjacency heatmap", marHeatmap = c(3,4,2,2),
                      plotDendrograms = FALSE, xLabelsAngle = 90)

```




<!-- Insert the session information into the document -->
```{r session-info}
```
